---
title: "Basic R Syntax and Data Types"
author: "Benjamin Hicks"
date: "2/27/2017"
output: md_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# R Syntax

R is derived from the [S programming language](https://en.wikipedia.org/wiki/S_(programming_language)),
which is an object oriented statistical language developed in the 1970s and
80s at Bell Labs by Rick Beker, Allan Wilks, and John Chambers.

As such, it tends to have more in common with other interpreted languages you 
may be familiar with (e.g., Python, Ruby) than other statistical programming
languages.

Many of R's operators will make instinctive sense to you though some may feel
strange.

## A simple R script

```{r}
a <- 1 
b <- 2
a+b
```

In this basic script, we set `a` to `1`, set `b` to `2` and then add them. You
could enter these commands at the R console and produce the same output.

The `<-` notation is probably new to you. R style tends to use this to indicate
that the output of a function, object, literal (i.e. 1, 'a string') is being
passed to another object.

Like in many programming languages, you can provide descriptive names to R variables
(which are naturally, objects).

```{r}
first_name <- 'Benjamin'
last_name <- 'Hicks'
cat(first_name, last_name, sep=' ')
```

In this example, I created two objects, then passed their value to the function
`cat` which concatenates them (using a space as a separator).

##  Data types
R is an implicitly-typed language. That means that a variable need not be declared
with a particular type. In C for example, you would have to create a variable
to hold an integer explicitly and that variable would remain that type.

R on the other hand, attempts to accomodate the data as best it can with implicit
conversions when needed. R has five basic classes of data: character (string),
numeric, integer (whole number), complex (numbers with i), and logical (TRUE/FALSE)


## Stupid vector tricks
All of these can be added to the most basic object: a vector. A vector is series
of objects all of the same class. This lets R do some interesting things without
resorting directly to arrays or other tricks that some programming languages require
to handle a series of data.

The `c` function lets you combine objects into a vector.
```{r}
a <- c(1, 2, 3)
b <- c(1, 2, 3)
a + b

```

`a` and `b` are both vectors of the numeric type. When I add them, R does something
you might not expect: it adds their constituent elements.

What happens when types are of multiple different data types? 

```{r}
a <- c(1, "aardvark", 3.4)
class(a)
a
```

R attempts to make a vector using the data type that can best 'hold' all the values.
In this case, "1" and "3.4" could at least be represented as characters to match
the character string "aardvar".

You can do this explicitly too using the `as` functions.

```{r}
a <- as.numeric("3.4")
class(a)
```

There is an exception to the rule that vectors must be of one type: lists.
These let you hold several different types of object in one 'vector'.

```{r}
test_list <- list('a', c(1,3), c(2,4))
# This is using subset notation to pick the second and third vectors of the
# list. Then a second notation to pick the first elements in each.
test_list[[2]][1] + test_list[[3]][1]

```

